Generalized Universe Hierarchies and First-Class Universe Levels

### 2021-02-20 paper.tex
progress

\newcommand{\Set}[1]{\mathsf{Set_{#1}}}
\newcommand{\Seti}{\mathsf{Set}}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\ext}{\triangleright}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
\newcommand{\mi}[1]{\mathit{#1}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\Id}{\mathsf{Id}}
\newcommand{\Nat}{\mathsf{Nat}}
\newcommand{\Bool}{\mathsf{Bool}}
\newcommand{\true}{\mathsf{true}}
\newcommand{\false}{\mathsf{false}}
\newcommand{\up}{\uparrow}
\newcommand{\down}{\downarrow}
\newcommand{\Lift}{\mathsf{Lift}}
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\Acc}{\mathsf{Acc}}
\newcommand{\acc}{\mathsf{acc}}
\newcommand{\Lvl}{\mathsf{Lvl}}
\renewcommand{\U}{\mathsf{U}}
\newcommand{\Code}{\mathsf{Code}}
\newcommand{\msf}[1]{\mathsf{#1}}
\newcommand{\uir}{\msf{U^{IR}}}
\newcommand{\elir}{\msf{El^{IR}}}
\newcommand{\ult}{\U_{<}}

@@

\begin{enumerate}
\item In Section \ref{sec:ttgu} we describe models of type theories where
  universe levels may come from any set with a well-founded transitive ordering
  relation. We specify models as categories equipped with level-indexed diagrams
  of families, as a variation on categories with families. Each morphism of
  levels is mapped to a lifting operation on terms and types. By varying the
  preservation properties of lifting operations, we can describe a range of
  stratification features, from two-level type theory to cumulative universes.
\item We use induction-recursion to model the mentioned theories. We model
  the strongest formulations for lifting and universes, namely cumulative
  universes with Russell-style type decoding.
\item In Section \ref{sec:ttfl} we describe type theories with internal types
  for levels and level morphisms, and extend the previous inductive-recursive
  semantics to cover these as well. Here, we can additionally represent various
  universe polymorphism features and level computations. We illustrate how
  predicative universe features in Coq and Agda can be justified in our framework.
\end{enumerate}

@@

\item
Besides consistency, we are often interested in \emph{canonicity},
\emph{normalization} or other metatheoretical properties. The current work
focuses on consistency and leaves other properties to future work. However, we
do keep canonicity in mind while specifying our systems, and aim to avoid
pitfalls which would cause canonicity to fail.
\item
We only focus on using universes as sized-based classifiers for
types. Universe-like stratification is also present in two-level type theory
\cite{twolevel}, modal type theories \cite{gratzer20multimodal} or as h-levels
in homotopy type theory \cite{hottbook}, but we do not attempt to cover such use
cases.
\item
We only cover predicative hierarchies.
\end{itemize}

@@

@@

\todo{ordinal indexed levels and lubs}

@@

\begin{mydefinition}[IR codes]
  Assume $i : \Lvl$ and $f : (j : \Lvl) \to j < i \to \Set0$. We define
  $\uir$ and $\elir$ by simultaneous IR. For basic type formers, below we only
  include codes for function types and Booleans; other type formers are
  straightforward to add (and we do have more in our Agda formalization).
\begin{alignat*}{5}
  & \uir   &&: \Set0
     && \rlap{$\elir : \uir \to \Set0$} \\
  & \U'    &&: (j : \Lvl) \to j < i \to \uir
     && \elir\,(\U'\,j\,p) &&= f\,j\,p \\
  & \Pi'   &&: (A : \uir) \to (\elir\,A \to \uir) \to \uir \hspace{2em}
     && \elir\,(\Pi'\,A\,B) &&= (a : \elir\,A) \to \elir\,(B\,a)\\
  & \Bool' &&: \uir
     && \elir\,\Bool'      &&= \Bool
\end{alignat*}
\end{mydefinition}

@@

($\uir$, $\elir$) can be viewed as a \emph{universe operator}: given semantics
for an initial segment of $\Lvl$ (given by $i$ and $f$), we create a new
universe which is closed under basic type formers, and also closed under all
sets in $f$ by the way of $\U'$. Most importantly, this operation can be
transfinitely iterated. We first define universes for initial segments of
$\Lvl$, by induction on the accessibility of levels:
\begin{alignat*}{3}
  & \ult\,: \{i : \Lvl\}\{p : \Acc\,<\,i\} \to (j : \Lvl) \to j < i \to \Set0 \\
  & \ult\,\{j\}\,\{\acc\,f\}\,j\,p = \uir_{j\,(\ult\,\{j\}\,\{f\,j\,p\})}
\end{alignat*}

\begin{mydefinition}[Semantic universe]
Since every level is accessible, we can define the full semantic hierarchy and
its decoding function.
\begin{alignat*}{3}
  &\U : \Lvl \to \Set0
     && \El : \{i : \Lvl\} \to \U\,i \to \Set0 \\
  &\U\,i = \uir_{i\,(\ult\,\{i\}\,\{\mathsf{<\!wf}\,i\})}\hspace{1em}
     && \El\,\{i\} = \elir_{i\,(\ult\,\{i\}\,\{\mathsf{<\!wf}\,i\})}
\end{alignat*}
\end{mydefinition}

\begin{mylemma}\label{lem:ucomp}
Assuming $p : i < j$, we have the computation rule $\ult\,j\,p = \U\,j$. Proof: we
may assume that any witness for $\Acc\,<\,i$ is of the form $\acc\,f$ for
some $f$. Then the equation becomes $\uir_{j\,(\ult\,\{j\}\,\{f\,j\,p\})} =
\uir_{j\,(\ult\,\{j\}\,\{\mathsf{<\!wf}\,j\})}$, but by Lemma \ref{lem:accprop}
the $f\,j\,p$ and $\mathsf{<\!wf}\,j$ witnesses are equal. \qed
\end{mylemma}

\begin{mydefinition}[Semantic $\Lift$]
We define by induction on $\uir$ a function with type $(p : i < j)(A : \U\,i)
\to (A' : \U\,j)\times(\El\,A' = \El\,A)$. However, for the sake of clarity, we
present this here as two (mutual) functions:
\begin{alignat*}{3}
  &\Lift\,      &&: (p : i < j) \to \U\,i \to \U\,j\\
  &\msf{ElLift} &&: (p : i < j)(A : \U\,i) \to \El\,(\Lift\,A) = \El\,A
\end{alignat*}
Let us look at $\Lift$ first:
\begin{alignat*}{3}
  &\Lift\,p\,(\U'\,k\,q)    &&= \U'\,k\,(p \circ q)\\
  &\Lift\,p\,(\Pi'\,A\,B)   &&= \Pi'\,(\Lift\,p\,A)\,(\lambda\,a.\,\Lift\,p\,(B\,a))\\
  &\Lift\,p\,\Bool'         &&= \Bool'
\end{alignat*}
Above, the $\Pi'$ definition is well-typed by $\msf{ElLift}\,p\,A$. For the
proof of $\msf{ElLift}$, the only interesting case is $\U'$. Here, we need
show $\ult\,\{j\}\,k\,(p \circ q) = \ult\,\{i\}\,k\,q$, but by Lemma
\ref{lem:ucomp} both sides are $\U\,k$. \qed
\end{mydefinition}

@@

\subsection{Model of TTGU}

@@

\begin{mydefinition}[Family diagram]
We map $i : \Lvl$ to a family structure as follows.
\begin{alignat*}{5}
  &\bm{\Ty}_i\,\Gamma = \Gamma \to \U\,i \hspace{2em} \bm{\Tm}_i\,\Gamma\,A = (\gamma : \Gamma) \to \El\,(A\,\gamma)
\end{alignat*}
Type and term substitution are given by composition with some $\sigma : \Gamma
\to \Delta$ function. Comprehension structure is given by $\Gamma \bm{\ext} A =
(\gamma : \Gamma) \times \El\,(A\,\gamma)$. Type lifting along $p : i < j$ is
as follows:n
\begin{alignat*}{5}
  & \bm{\Lift}_{i}^{j}p : \bm{\Ty}_i\,\Gamma \to \bm{\Ty}_j\,\Gamma\\
  & \bm{\Lift}_{i}^{j}p\,A = \lambda\,\gamma.\,\Lift_{i}^{j}p\,(A\,\gamma)
\end{alignat*}
Now, two of the strict inclusion equations follow from $\msf{ElLift}$, namely
$(\Gamma \bm{\ext} \bm{\Lift}_{i}^{j}p\,A) = (\Gamma \bm{\ext} A)$ and
$\bm{\Tm}_j\,\Gamma\,(\bm{\Lift}_{i}^{j}p\,A) = \bm{\Tm}_i\,\Gamma\,A$. Thus, we
can just define term lifting as $\bm{\up}_{i}^{j}\!p\,t = t$ and
$\bm{\down}_{i}^{j}\!p\,t = t$. Basic type formers are as follows.
\begin{alignat*}{5}
  & \bm{\Pi}\,A\,B = \lambda\,\gamma.\,\Pi'\,(A\,\Gamma)\,(\lambda\,\alpha.\,B\,(\gamma,\,\alpha))
  & \hspace{2em}\bm{\Bool} = \lambda\,\gamma.\,\Bool'
\end{alignat*}
$\bm{\Lift}_{i}^{j}p$ preserves type formers by Lemma \ref{lem:liftprop}. We
define basic term formers using metatheoretic term formers, e.g.\ $\bm{\true}_i
= \lambda\,\gamma.\,\true$ and $(\bm{\lambda}_i\,x.\,t) =
\lambda\,\gamma\,\alpha.\,t\,(\gamma,\,\alpha)$. Note that since semantic term
formers are just external constructors, they do not depend on levels, so
e.g.\ $\bm{\true}_i$ is the same at all $i$. This implies that
$\bm{\up}_{i}^{j}\!p$ preserves term formers as well, so ($\bm{\Lift}_{i}^{j}p$,
$\bm{\up}_{i}^{j}\!p$, $\bm{\down}_{i}^{j}\!p$) is a strict family inclusion.

@@

%% To verify that we have strict inclusions, we need to
%% show that $\bm{\Lift}$ and $\bm{\up}$ preserve all type and term
%% formers, but first we look at the semantics of type and term formers.

%% To verify that we have strict inclusions, it
%% remains to show that $\bm{\Lift}$ and $\bm{\up}$ preserve all type and term
%% formers.  For $\bm{\Lift}$ this follows from Lemma \ref{lem:liftprop}. For
%% $\bm{\up}$, note that semantics terms

  %% & (\Gamma \ext \Lift\,A) &&= (\Gamma \ext A)     \label{eq:cumcon}    \\
  %% & \Tm_1\,\Gamma\,(\Lift\,A) &&= \Tm_0\,\Gamma\,A  \label{eq:cumtm}     \\
  %% & \up\!t &&= t                                    \label{eq:cumlift}

%% The strict inclusions are given as follows.
  %% Lift : ∀ {Γ i j} → i < j → Ty Γ i → Ty Γ j
  %% Lift p A γ = ↑ p (A γ)

@@

Specification of the theory

>>>>

\newcommand{\Set}[1]{\mathsf{Set_{#1}}}
\newcommand{\Seti}{\mathsf{Set}}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\id}{\mathsf{id}}
\newcommand{\ext}{\triangleright}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
\newcommand{\mi}[1]{\mathit{#1}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\Id}{\mathsf{Id}}
\newcommand{\Nat}{\mathsf{Nat}}
\newcommand{\Bool}{\mathsf{Bool}}
\newcommand{\true}{\mathsf{true}}
\newcommand{\false}{\mathsf{false}}
\newcommand{\up}{\uparrow}
\newcommand{\down}{\downarrow}
\newcommand{\Lift}{\mathsf{Lift}}
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\Acc}{\mathsf{Acc}}
\newcommand{\acc}{\mathsf{acc}}
\newcommand{\Lvl}{\mathsf{Lvl}}
\renewcommand{\U}{\mathsf{U}}
\newcommand{\Code}{\mathsf{Code}}
\newcommand{\msf}[1]{\mathsf{#1}}
\newcommand{\uir}{\msf{U^{IR}}}
\newcommand{\elir}{\msf{El^{IR}}}
\newcommand{\ult}{\U_{<}}
\newcommand{\mkMor}{\msf{mk}\!_<}
\newcommand{\unMor}{\msf{un}\!_<}
\newcommand{\mkLvl}{\msf{mk}_{\Lvl}}
\newcommand{\unLvl}{\msf{un}_{\Lvl}}

@@

\begin{enumerate}
\item In Section \ref{sec:ttgu} we describe models of type theories where
  universe levels may come from any set with a well-founded transitive ordering
  relation. We specify models as categories equipped with level-indexed diagrams
  of families, as a variation on categories with families. Each morphism of
  levels is mapped to a lifting operation on terms and types. By varying the
  preservation properties of lifting operations, we can describe a range of
  stratification features, from two-level type theory to cumulative universes.
\item We use induction-recursion to model the mentioned theories. We model
  the strongest formulations for lifting and universes, namely cumulative
  universes with Russell-style type decoding.
\item In Section \ref{sec:ttfl} we describe type theories with internal types
  for levels and level morphisms, and extend the previous inductive-recursive
  semantics to cover these as well. Here, we can additionally represent various
  universe polymorphism features and level computations.
\end{enumerate}

@@

\item
Besides consistency, we are often interested in \emph{canonicity},
\emph{normalization} or other metatheoretical properties. The current work
focuses on consistency and leaves other properties to future work. However, we
do keep canonicity in mind while specifying our systems, and aim to avoid
pitfalls which would cause canonicity to fail.
\item
We only focus on using universes as sized-based classifiers for
types. Universe-like stratification is also present in two-level type theory
\cite{twolevel}, modal type theories \cite{gratzer20multimodal} or as h-levels
in homotopy type theory \cite{hottbook}, but we do not attempt to cover such use
cases.
\item
We only cover predicative hierarchies.
\item
We do not discuss ergonomics and implementation of elaboration. Our goal is to
develop a system which is as flexible as possible, even though that may not be
the best suited for inference and elaboration. For example, first-class levels
have undecidable ordering in general.
\end{itemize}

@@

\todo{purge unused features}

@@

@@

\begin{mydefinition}[IR codes]
  Assume $i : \Lvl$ and $f : (j : \Lvl) \to j < i \to \Set0$. We define $\uir$
  and $\elir$ by simultaneous IR. For basic type formers, below we only include
  codes for function types, the empty type, and $\Bool$; other type formers are
  straightforward to add (and we do have more in the Agda formalization).
\begin{alignat*}{5}
  & \uir   &&: \Set0
     && \rlap{$\elir : \uir \to \Set0$} \\
  & \U'    &&: (j : \Lvl) \to j < i \to \uir
     && \elir\,(\U'\,j\,p) &&= f\,j\,p \\
  & \Pi'   &&: (A : \uir) \to (\elir\,A \to \uir) \to \uir \hspace{2em}
     && \elir\,(\Pi'\,A\,B) &&= (a : \elir\,A) \to \elir\,(B\,a)\\
  & \bot' &&: \uir
     && \elir\,\bot'      &&= \bot\\
  & \Bool' &&: \uir
     && \elir\,\Bool'     &&= \Bool
\end{alignat*}
\end{mydefinition}

@@

($\uir$, $\elir$) can be viewed as a \emph{universe operator}: given semantics
for an initial segment of $\Lvl$ (given by $i$ and $f$), we create a new
universe which is closed under basic type formers, and also closed under all
sets in $f$ by the way of $\U'$. Most importantly, this operation can be
transfinitely iterated. We first define universes for initial segments of
$\Lvl$, by induction on the accessibility of levels:
\begin{alignat*}{3}
  & \ult\,: (i : \Lvl)\{p : \Acc\,<\,i\} \to (j : \Lvl) \to j < i \to \Set0 \\
  & \ult\,i\,\{\acc\,f\}\,j\,p = \uir_{j\,(\ult\,j\,\{f\,j\,p\})}
\end{alignat*}

\begin{mydefinition}[Semantic universe]
Since every level is accessible, we can define the full semantic hierarchy and
its decoding function.
\begin{alignat*}{3}
  &\U : \Lvl \to \Set0
     && \El : \{i : \Lvl\} \to \U\,i \to \Set0 \\
  &\U\,i = \uir_{i\,(\ult\,i\,\{\mathsf{<\!wf}\,i\})}\hspace{1em}
     && \El\,\{i\} = \elir_{i\,(\ult\,i\,\{\mathsf{<\!wf}\,i\})}
\end{alignat*}
\end{mydefinition}

\begin{mylemma}\label{lem:ucomp}
Assuming $p : i < j$, we have the computation rule $\ult\,j\,p = \U\,j$. Proof: we
may assume that any witness for $\Acc\,<\,i$ is of the form $\acc\,f$ for
some $f$. Then the equation becomes $\uir_{j\,(\ult\,j\,\{f\,j\,p\})} =
\uir_{j\,(\ult\,j\,\{\mathsf{<\!wf}\,j\})}$, but by Lemma \ref{lem:accprop}
the $f\,j\,p$ and $\mathsf{<\!wf}\,j$ witnesses are equal. \qed
\end{mylemma}

\begin{mydefinition}[Semantic $\Lift$]
We define by induction on $\uir$ a function with type $(p : i < j)(A : \U\,i)
\to (A' : \U\,j)\times(\El\,A' = \El\,A)$. However, for the sake of clarity, we
present this here as two (mutual) functions:
\begin{alignat*}{3}
  &\Lift\,      &&: (p : i < j) \to \U\,i \to \U\,j\\
  &\msf{ElLift} &&: (p : i < j)(A : \U\,i) \to \El\,(\Lift\,A) = \El\,A
\end{alignat*}
Let us look at $\Lift$ first:
\begin{alignat*}{3}
  &\Lift\,p\,(\U'\,k\,q)    &&= \U'\,k\,(p \circ q)\\
  &\Lift\,p\,(\Pi'\,A\,B)   &&= \Pi'\,(\Lift\,p\,A)\,(\lambda\,a.\,\Lift\,p\,(B\,a))\\
  &\Lift\,p\,\bot'          &&= \bot'\\
  &\Lift\,p\,\Bool'         &&= \Bool'
\end{alignat*}
Above, the $\Pi'$ definition is well-typed by $\msf{ElLift}\,p\,A$. For the
proof of $\msf{ElLift}$, the only interesting case is $\U'$. Here, we need
show $\ult\,j\,k\,(p \circ q) = \ult\,i\,k\,q$, but by Lemma
\ref{lem:ucomp} both sides are $\U\,k$. \qed
\end{mydefinition}

@@

\subsection{IR Model of TTGU}

@@

\begin{mydefinition}[Family diagram]
We map $i : \Lvl$ to a family structure as follows.
\begin{alignat*}{5}
  &\bm{\Ty}_i\,\Gamma = \Gamma \to \U\,i \hspace{2em} \bm{\Tm}_i\,\Gamma\,A = (\gamma : \Gamma) \to \El\,(A\,\gamma)
\end{alignat*}
Type and term substitution are given by composition with some $\sigma : \Gamma
\to \Delta$ function. Comprehension structure is given by $\Gamma \bm{\ext} A =
(\gamma : \Gamma) \times \El\,(A\,\gamma)$. Type lifting along $p : i < j$ is
as follows:
\begin{alignat*}{5}
  & \bm{\Lift}_{i}^{j}p : \bm{\Ty}_i\,\Gamma \to \bm{\Ty}_j\,\Gamma\\
  & \bm{\Lift}_{i}^{j}p\,A = \lambda\,\gamma.\,\Lift_{i}^{j}p\,(A\,\gamma)
\end{alignat*}
Now, two of the strict inclusion equations follow from $\msf{ElLift}$, namely
$(\Gamma \bm{\ext} \bm{\Lift}_{i}^{j}p\,A) = (\Gamma \bm{\ext} A)$ and
$\bm{\Tm}_j\,\Gamma\,(\bm{\Lift}_{i}^{j}p\,A) = \bm{\Tm}_i\,\Gamma\,A$. Thus, we
can just define term lifting as $\bm{\up}_{i}^{j}\!p\,t = t$ and
$\bm{\down}_{i}^{j}\!p\,t = t$. Basic type formers are as follows.
\begin{alignat*}{5}
  & \bm{\Pi}\,A\,B = \lambda\,\gamma.\,\Pi'\,(A\,\Gamma)\,(\lambda\,\alpha.\,B\,(\gamma,\,\alpha))
  & \hspace{2em}\bm{\bot}_i = \lambda\,\gamma.\,\bot'
  & \hspace{2em}\bm{\Bool}_i = \lambda\,\gamma.\,\Bool'
\end{alignat*}
$\bm{\Lift}_{i}^{j}p$ preserves type formers by Lemma \ref{lem:liftprop}. We
define basic term formers and eliminators using metatheoretic features, e.g.\ $\bm{\true}_i
= \lambda\,\gamma.\,\true$ and $(\bm{\lambda}_i\,x.\,t) =
\lambda\,\gamma\,\alpha.\,t\,(\gamma,\,\alpha)$. Note that since semantic term
formers are just external constructors, they do not depend on levels, so
e.g.\ $\bm{\true}_i$ is the same at all $i$. This implies that
$\bm{\up}_{i}^{j}\!p$ preserves term formers as well, so ($\bm{\Lift}_{i}^{j}p$,
$\bm{\up}_{i}^{j}\!p$, $\bm{\down}_{i}^{j}\!p$) is a strict family inclusion.

@@

@@

In the following, we specify and model type theories where levels and
their morphisms are represented by internal types.

However, it would be awkward to pick a particular structure for levels, and
specify a type theory which internalizes that structure; for example
internalizing levels as natural numbers. We do not want to repeat the
specification and semantics for each choice of level structure; instead, we aim
to have a more generic solution.
\begin{enumerate}
  \item We first give a specification of \emph{type theory with dependent
        levels}, or TTDL, where levels and level morphism may depend on typing contexts.
        Here, liftings, universes and type formers are specified, but the internal
        structure of levels is not yet pinned down.
  \item We show that we can extend TTDL with \emph{level reflection} rules, which identify
        levels with particular internal types, thereby getting \emph{type theories with first-class
        levels}, or TTFL.
\end{enumerate}
This decreases the amount of work that we have to do, in order to get semantics
for different level setups. We only need to pick an external level structure
such that it can be also represented using TTDL type formers.

\begin{mydefinition} A model of TTDL consists of the following.
\begin{enumerate}
\item
  A base category ($\Con$, $\Sub$) with terminal object $\emptycon$.
\item
  A ``dependent'' level structure on the base category:
  \begin{alignat*}{3}
    &\Lvl                  &&: \Con \to \Seti\\
    &\blank\!<\!\blank     &&: \{\Gamma : \Con\} \to \Lvl\,\Gamma \to \Lvl\,\Gamma \to \Seti\\
    &<\!\msf{prop}         &&: (p\,q : i < j) \to p = q\\
    &\blank\!\circ\!\blank &&: j < k \to i < j \to i < k
  \end{alignat*}
  Additionally, $\Lvl$ and $\blank\!<\!\blank$ are natural in the base category,
  so they support substitution operations. \emph{Remark:} at this point, we do not require well-foundedness
  for $\blank\!<\!\blank$, as it has no bearing on basic lifting and universe rules, and
  well-foundedness will be usually internally provable when we add level reflection rules.
\item
  A ``bootstrapping'' assumption on levels. This can be any collection of levels
  and morphisms. We pick the assumption that $l_0, l_1 : \Lvl\,\Gamma$ exist
  together with $l_{01} : l_0 < l_1$. We need to assume at least some levels,
  because otherwise the TTDL syntax is trivial; it is not even possible to
  write down a non-empty context or a closed type former. $l_0 < l_1$ allows large
  eliminations on type formers so it provides a fair amount of expressive power.
\item
  A family structure:
  \begin{alignat*}{3}
    &\Ty &&: (\Gamma : \Con) \to \Lvl\,\Gamma \to \Seti\\
    &\Tm &&: (\Gamma : \Con)\{i : \Lvl\,\Gamma\} \to \Ty\,\Gamma\,i \to \Seti\\
    &\blank\ext\blank &&: (\Gamma : \Con)\{i : \Lvl\,\Gamma\} \to \Ty\,\Gamma\,i \to \Con
  \end{alignat*}
  We have type and term substitution, which depends on level substitution. For instance, we have
  \[
    \blank[\blank] : \Ty\,\Delta\,i \to (\sigma : \Sub\,\Gamma\,\Delta) \to \Ty\,\Gamma\,(i[\sigma])
  \]
  We also have a comprehension isomorphism $\Sub\,\Gamma\,(\Delta \ext A) \simeq
  (\sigma : \Sub\,\Gamma\,\Delta)\times \Tm\,\Gamma\,(A[\sigma])$, which is
  natural in $\Gamma$.
\item A lifting structure with
  \begin{alignat*}{3}
    &\Lift &&: \{\Gamma : \Con\}\{i\,j : \Lvl\,\Gamma\} \to i < j \to \Ty\,\Gamma\,i \to \Ty\,\Gamma\,j\\
    &\up   &&: \{\Gamma : \Con\}\{i\,j : \Lvl\,\Gamma\}(p : i < j) \to \Tm\,\Gamma\,A \to \Tm\,\Gamma\,(\Lift\,p\,A)
  \end{alignat*}
  Such that
  \vspace{1em}
  \begin{enumerate}
    \item $\Lift$ preserves all basic type formers and has functorial action on $p \circ q$.
    \item $\up$ has an inverse $\down$, preserves all basic term formers and has functorial action on $p \circ q$.
    \item $(\Gamma \ext A) = (\Gamma \ext \Lift\,p\,A)$, and $\Tm\,\Gamma\,A = \Tm\,\Gamma\,(\Lift\,p\,A)$ and $\up\,t = t$.
    %% \item $\Tm\,\Gamma\,A = \Tm\,\Gamma\,(\Lift\,p\,A)$.
    %% \item $\up\,t = t$.
  \end{enumerate}
  \vspace{1em}
  Above we mention basic type formers, although we haven't yet specified
  those. The way this should be understood, is that any basic type former introduced
  from now on should come equipped with preservation equations for lifting. This is
  similar to how we mandate that any introduced type former must be natural with respect
  to substitution.
\item A universe structure
  \begin{alignat*}{6}
    &\U  &&: \{\Gamma : \Con\}(i\,j : \Lvl\,\Gamma) \to i < j \to \Ty\,\Gamma\,j\hspace{2em}
    &\El &&: \Tm\,\Gamma\,(\U\,i\,j\,p) \to \Ty\,\Gamma\,i
  \end{alignat*}
  Such that $\Lift\,p\,(\U\,i\,j\,q) = \U\,i\,k\,(p \circ q)$, $\El$ has
  inverse $\Code$, $\Tm\,\Gamma\,(\U\,i\,j\,p) = \Ty\,\Gamma\,i$ and $\El\,t =
  t$.
\item Basic type formers.
\end{enumerate}
\end{mydefinition}

\begin{mydefinition}[IR model of TTDL]\label{def:ttdlmodel}
 Assume an external $\Lvl$ structure such that it supports $l_0, l_1 : \Lvl$ and
 $l_{01} : l_0 < l_1$ (the bootstrapping assumption). We again use the IR
 universe constructions from Section \ref{sec:inductive_recursive_codes},
 instantiated to the assumed $\Lvl$ structure. We describe components of the
 model in order. Again, we write components of the model in \textbf{bold} font.
\begin{enumerate}
\item The base category remains unchanged from the TTGU model.
\item For the level structure, we define $\bm{\Lvl}\,\Gamma = \Gamma \to \Lvl$ and
  $i \bm{<} j = (\gamma : \Gamma) \to i\,\gamma < j\,\gamma$. Subsitution for internal
  levels and morphisms is given by function composition with $\sigma : \Gamma \to \Delta$.
  Internal composition and $\bm{<\!\msf{prop}}$ follow from the external counterparts.
\item The internal bootstrapping assumption is modeled with the external counterpart.
\item We define $\bm{\Ty}\,\Gamma\,i = (\gamma : \Gamma) \to \U\,(i\,\gamma)$ and
  $\bm{\Tm}\,\Gamma\,A = (\gamma : \Gamma) \to \El\,(A\,\gamma)$. Substitution is
  again function composition, and we have $\Gamma \bm{\ext} A = (\gamma :
  \Gamma)\times \El\,(A\,\gamma)$.
\item Type lifting is given by $\bm{\Lift}\,p\,A = \lambda\,\gamma.\,
  \Lift\,(p\,\gamma)\,(A\,\gamma)$.  Similarly as in the TTGU model,
  $\bm{\Tm}\,\Gamma\,A = \bm{\Tm}\,\Gamma\,(\bm{\Lift}\,p\,A)$ and $\Gamma \bm{\ext}
  A = \Gamma \bm{\ext} (\bm{\Lift}\,p\,A)$ follow from the $\msf{ElLift}$
  equality, and term lifting is the identity function.
\item We define $\bm{\U}\,i\,j\,p = \lambda\,\gamma.\,\U'\,(p\,\gamma)$. Again,
  we have $\bm{\Tm}\,\Gamma\,(\U\,i\,j\,p) = \bm{\Ty}\,\Gamma\,i$ by Lemma \ref{lem:ucomp},
  and $\bm{\El}$ and $\bm{\Code}$ are identity functions.
\item Basic type formers are interpreted using $\uir$ codes. Preservation of
  type and term formers by lifting follows by the definition of $\Lift$ and
  $\El$.
\end{enumerate}
\end{mydefinition}

To summarize, the only interesting change compared to the TTGU model is that
levels and level morphisms gain potential dependency on contexts. However, in the
IR model, this is simply the addition of an extra semantic function parameter.

\subsection{Level Reflection}\label{sec:level_reflection}

\begin{mydefinition}[Level reflection rules]
Assume that we have \emph{definitions for internal levels} in the syntax of TTDL (i.e.\
the initial TTDL model):
\begin{alignat*}{3}
  &\Lvl^I               &&: \Ty\,\Gamma\,l_0\\
  &l_0^I,\,l_1^I         &&: \Tm\,\Gamma\,\Lvl^I \\
  &\blank\!<^I\!\blank  &&: \Tm\,\Gamma\,\Lvl^I \to \Tm\,\Gamma\,\Lvl^I \to \Ty\,\Gamma\,l_0\\
  &l_{01}^{I}            &&: \Tm\,\Gamma\,(l_0^I <^I l_1^I)
\end{alignat*}
A \emph{reflection rule} for the above consists of
\begin{enumerate}
  \item $\mkLvl : \Tm\,\Gamma\,\Lvl^I \to \Lvl\,\Gamma$ with its inverse $\unLvl$, such that $\mkLvl\,l_0^I = l_0$ and $\mkLvl\,l_1^I = l_1$.
  \item $\mkMor\,\,: \Tm\,\Gamma\,(i <^I j) \to \mkLvl\,i < \mkLvl\,i$ with its inverse
        $\unMor$.
\end{enumerate}
\end{mydefinition}

For any definition of internal levels, we may extend the specification of TTDL
with the corresponding reflection rule, thereby getting an algebraic signature
for a TTFL. We can easily get a TTFL with an IR model in the following
way. First, pick an external $\Lvl$ structure which a) satisfies the
bootstrapping assumption b) has sets of levels and morphisms which can be
represented with syntactic TTDL types.

For example, if $\Lvl$ = ($\Nat$, $\blank\!<\!\blank$), with $l_0 = 0$ and $l_1
= 1$, and TTDL supports natural numbers, then we can define $\Lvl^I$ as the
internal $\Nat_{l_0}$, and define $\blank\!<^I\!\blank$ as the usual ordering of
numbers, using TTDL type formers and large elimination (which is available from
$l_0 < l_1$). Then it follows that the model in Definition \ref{def:ttdlmodel},
instantiated to the current level structure, satisfies level reflection. The
model even supports the stricter $\Tm\,\Gamma\,\Nat_{l_0} = \Lvl\,\Gamma$
equation, but in general it is easier to set up models if only an isomorphism
is required.

\section{Universe Features in TTFL}

In this section we give a tour of some of the features expressible in TTFL.

\textbf{Bounded universe polymorphism} is realized by quantifying
over levels and morphisms with the usual $\Pi$ types. For example, if internal
levels strictly correspond to natural numbers, we may have
\begin{alignat*}{3}
  &\msf{idUpTo3} : \Pi(l : \Nat_3)(p : l <^I 3)(A : \U\,l\,3\,(\mkMor\,p)) \to \Lift\,(\mkMor\,p)\,A \to \Lift\,(\mkMor\,p)\,A\\
  &\msf{idUpTo3} = \lambda\,l\,p\,A\,a.\,a
\end{alignat*}
Here, we make sure that all types in the $\Pi$ are on the same level. We assume
that internal levels are in $\Nat_0$, but we can bind an $l : \Nat_3$, because
by cumulativity $l$ is also a term of $\Nat_0$.

\textbf{Transfinite hierarchies} are naturally supported. For example, $\Lvl$
can be identified with $\msf{Maybe}\,\Nat_0$, where $\msf{Nothing}$ defines
$\omega$ and $\msf{Just}\,n$ is a finite level. Then, by the definition of
morphisms, we have $<\!\omega : \Pi(n : \Nat_0) \to \msf{Just}\,n <^I \omega$.
We can use this to quantify over finite levels, as in the following type:
\begin{alignat*}{3}
  \Pi(n : \Nat_{\omega})(A : \U\,n\,\omega\,(\mkMor\,(<\!\omega\,n))) \to \Lift\,(\mkMor\,(<\!\omega\,n))\,A \to \Lift\,(\mkMor\,(<\!\omega\,n))\,A
\end{alignat*}

\textbf{Induction on levels and level morphisms.} In Agda 2.6.1, there is an
internal type of finite levels, and while construction rules and some built-in
operations on levels are exposed, there is no general elimination rule on levels.
Thus, there is a $\Nat \to \Lvl$ function but no non-constant $\Lvl \to \Nat$ one.
In contrast, TTFL supports arbitrary elimination on levels, and in particular levels
can be definitionally isomorphic to natural numbers.

\textbf{Type formers returning in least upper bounds of levels}. It is common in
type theories to allow type formers to have parameter types in different
universe levels, say $i$ and $j$, and return in level $i \sqcup j$. In TTFL,
whenever levels are \emph{trichotomous}, meaning that the ordering and equality of
levels is internally decidable, then $\sqcup$ is a definable operation, and the
``heterogeneous'' type formers are derivable\footnote{A level structure which is trichotonomous and supports \emph{extensionality}, i.e.\ $(\forall i.\, (i < j) \iff (i < k)) \to j = k$, is called a \emph{type-theoretic ordinal}. Assuming excluded middle type-theoretic ordinals are equivalent to classical ordinals \cite[Section 10.3]{hottbook}.}.

\textbf{Coercive cumulative subtyping.} TTFL does not directly support
subtyping. However, given enough type formers, it is possible to simply define
subtyping as an interal inductive relation, and interpret a witness of $A \leq B$ as
a coercion function from $A$ to $B$. Also, it should be possible to model a
theory with cumulative subtyping by translation to coercions in TTFL.
